Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-10 appendix.tex
nearly done with major update
paper updates, code updates
added new appendix - repeat gordon's notes with iso eos

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}), which states that
instability in the adiabatic limit requires the disk to be close to
neutrally stratified. %also need real freq^2 in adia limit 
This is because the VGSB retains the destabilizing effect of vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial disk
struture responsible for it. Nevertheless, we demonstrate below that
this inconsistency is unimportant for the vertical shear instability,
which occurs for $\beta\ll1$.

@@

\section{Full governing equation for vertically isothermal disks}\label{adia_improve}
% \section{Improving the nearly-Keplerian approximation for adiabatic
%   disks}\label{adia_improve}
In \S\ref{approx_gov} we made the replacement $D\to\Omega_k^2$ before
eliminating variables to obtain a single equation for $\delta v_z$,
Eq. \ref{vertiso_gov}.  This procedure ignores the vertical dependence of
$D=\kappa^2(z) - \sigma^2$. We show here that this has no significant
consequence for thin disks.

@@

 Eliminating $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
 Eq. \ref{lin_vz} for vertically isothermal disks, retaining the
 vertical dependence of $D$ and including thermal relaxation, we
 obtain 
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low-frequency approximation with the
replacement $D\to \Omega_k^2$ gives, in terms of dimensionless
variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
neglecting these terms has no qualitative effect on the discussion in
\S\ref{analytic_adia} and \S\ref{analytic_relax}.  This issue does not
arise for isothermal perturbations discussed in \S\ref{iso_discuss}
since in that case we work with a governing equation for $W$ instead.

@@

\section{Coefficients for the dispersion relation for perturbations
with thermal relaxation}\label{relax_coeff}
The coefficients of the dispersion relation Eq. \ref{relax_disp} is
given by:
\begin{align}
  &c_0 = M(M+1)\widetilde{A}^2,\\
  &c_1 = \ii\beta\left\{\left(1-\gamma\right)\left[1 +
      \khat^2\left(1+2M\right)^2 - 4 \ii\epsilon q\khat M (M+1)\right] 
    - 2\widetilde{A}^2\gamma M (M+1)\right\},\\
  &c_2 = \left(\khat^2 + 1\right)\widetilde{A} + \beta^2\left\{(1-\gamma)\left[1
      + \gamma \khat^2(1+2M)^2 - 4\ii\epsilon q \khat \gamma M(M+1)
    \right]
    -\gamma^2 \widetilde{A}^2 M(M+1)
  \right\},\\
  &c_3 = \beta\left\{\epsilon q \khat + \gamma \left[\ii + \epsilon q
      k \left(1+2\khat^2\right)\right] - 3\ii - 2\ii
    \khat^2\right\},\\
  &c_4 =
  \beta^2\left(1+\gamma\khat^2\right)\left[\gamma\left(1-\ii\epsilon q
    \khat\right)-2\right] - \left(1+\khat^2\right)^2,\\
&c_5 = 2\ii\beta\left(1+\khat^2\right)\left(1+\gamma\khat^2\right),\\
&c_6 = \beta^2\left(1+\gamma\khat^2\right)^2.
\end{align}

>>>>

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}), which states that
instability in the adiabatic limit requires the disk to be close to
neutrally stratified. %also need real freq^2 in adia limit 
This is because the VGSB retains the destabilizing effect of vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial disk
structure responsible for it. Nevertheless, we demonstrate below that
this inconsistency is unimportant for the vertical shear instability,
which occurs for $\beta\ll1$.

@@

\section{Full governing equation for vertically isothermal disks in
  the VGSB}\label{adia_improve}
% \section{Improving the nearly-Keplerian approximation for adiabatic
%   disks}\label{adia_improve}
In \S\ref{approx_gov} we made the replacement $D\to\Omega_k^2$ before
eliminating variables to obtain a single equation for $\delta v_z$,
Eq. \ref{vertiso_gov}.  This procedure ignores the vertical dependence of
$D=\kappa^2(z) - \sigma^2$. We show here that this has no significant
consequence for thin disks.

@@

 Eliminating $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
 Eq. \ref{lin_vz} for vertically isothermal disks, retaining the
 vertical dependence of $D$ and including thermal relaxation, we
 obtain 
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low-frequency approximation with the
replacement $D\to \Omega_k^2$ gives, in terms of dimensionless
variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
neglecting these terms has no qualitative effect on the discussion in 
\S\ref{analytic_relax}.  This issue does not
arise for isothermal perturbations discussed in \S\ref{iso_discuss}
since in that case we work with a governing equation for $W$ instead.

@@

\section{Eigenfrequency in locally
  isothermal disks in global cylindrical geometry}
Consider a locally isothermal disk in 3D such that $P=c_s^2\rho$ at
all times, so in the perturbed state $\delta P =
c_s^2\delta\rho$, with $c_s(r)$ begin a prescribed function. 
The linearized fluid equations in cylindrical
geometry is then 
\begin{align}
 \ii\sigma \delta v_r + 2\Omega\delta v_\phi &= c_s^2\frac{\p}{\p r}\left(\frac{\delta\rho}{\rho}\right),\\
 \ii\sigma \delta v_\phi &= \frac{1}{r}\delta\bm{v}\cdot\nabla j,\\
 \ii\sigma \delta v_z &= c_s^2\frac{\p}{\p
   z}\left(\frac{\delta\rho}{\rho}\right),\\
 \ii\sigma\delta\rho &= \nabla\cdot\left(\rho\delta\bm{v}\right).
\end{align}
We can eliminate $\delta v_\phi$ from these equations and obtain
\begin{align}
  -\sigma^2\rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) +
  \kappa^2\rho|\delta v_r|^2 + \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
  \delta v_r^*\delta v_z =
  c_s^2\rho\delta\bm{v}^*\cdot\nabla\left(\frac{\ii\sigma\delta\rho}{\rho}\right)
  =c_s^2\rho\delta\bm{v}^*\cdot\nabla\left[\frac{1}{\rho}\nabla\cdot\left(\rho\delta\bm{v}\right)\right].\label{global_integral}
\end{align}
At this point if we make the anelastic approximation by assuming
$\nabla\cdot\left(\rho\delta\bm{v}\right)$ is negligible in this equation \citep{nelson13},
then the only contribution to instability comes from the third term on
the left hand side, i.e. the explicit vertical shear term $\p_zj$. The 
term proportional to $\kappa^2$ is stabilizing. Because radial shear
is much larger than vertical shear, this implies that significant
instability can only be achieved with large vertical
motions (i.e. $|\delta v_z|\gg|\delta v_r|$).

If we integrate Eq. \ref{global_integral} over a volume
and assume boundary terms vanish when integrating by parts, we obtain 
\begin{align}
  -\sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV +
  \int \kappa^2\rho|\delta v_r|^2 dV + \int \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
  \delta v_r^*\delta v_z dV = -\int
  \frac{c_s^2}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
  dV - \int \nabla\cdot\left(\rho\delta\bm{v}\right) \delta
  \bm{v}^*\cdot\nabla c_s^2.
\end{align}
Recalling $j=r^2\Omega$ and $r\p_z\Omega^2 = -\p_z\ln{\rho}\p_r c_s^2$
for a locally isothermal disk, we obtain
\begin{align}
  \sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV =
  \int \kappa^2\rho|\delta v_r|^2 dV + \int
  \frac{c_s^2}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
  dV + \int \left[ \rho \left(\nabla\cdot\delta\bm{v}\right)\delta
    v_r^* + |\delta v_r|^2\frac{\p \rho}{\p r}\right]\frac{dc_s^2}{dr}  dV.
\end{align}
We see that instability is only possible if
$dc_s^2/dr\neq0$. Furthermore, since density and temperature typically
decrease outwards in astrophysical disks, instability is associated
with the term proportional to $\nabla\cdot\delta\bm{v}$. It is also
clear that $\sigma^2$ will be complex in general.

\section{Coefficients for the dispersion relation for perturbations
with thermal relaxation}\label{relax_coeff}
The coefficients of the dispersion relation Eq. \ref{relax_disp} is
given by:
\begin{align}
  &c_0 = M(M+1)A^2,\\
  &c_1 = \ii\beta\left\{\left(1-\gamma\right)\left[1 +
      \khat^2\left(1+2M\right)^2 - 4 \ii\epsilon q\khat M (M+1)\right] 
    - 2A^2\gamma M (M+1)\right\},\\
  &c_2 = \left(\khat^2 + 1\right)A + \beta^2\left\{(1-\gamma)\left[1
      + \gamma \khat^2(1+2M)^2 - 4\ii\epsilon q \khat \gamma M(M+1)
    \right]
    -\gamma^2 A^2 M(M+1)
  \right\},\\
  &c_3 = \beta\left\{\epsilon q \khat + \gamma \left[\ii + \epsilon q
      k \left(1+2\khat^2\right)\right] - 3\ii - 2\ii
    \khat^2\right\},\\
  &c_4 =
  \beta^2\left(1+\gamma\khat^2\right)\left[\gamma\left(1-\ii\epsilon q
    \khat\right)-2\right] - \left(1+\khat^2\right)^2,\\
&c_5 = 2\ii\beta\left(1+\khat^2\right)\left(1+\gamma\khat^2\right),\\
&c_6 = \beta^2\left(1+\gamma\khat^2\right)^2.
\end{align}
